# Cleaning Code

This chapter walks you through all of the R code used to clean the raw data inputs. The resulting cleaned datasets are then used in the modeling (see next Chapter).

## Create Final Data Inputs

This script imports all of raw data inputs needed for the model.

Note that you must first open the 'methods_manual' R project before running this script or else it will not work.

First, let's set up our environment. 

```{r, results='hide', warning=FALSE, message=FALSE}

rm(list = ls())

options(scipen=999)

library(tidyverse)
library(readxl)
library(stringr)

# check working directory
getwd()

```

Create a date string that will be appended to all data file names.

```{r, results='hide', warning=FALSE, message=FALSE}

my_date <- paste0("_", Sys.Date(), "_FINAL.csv")

```

### Other Data Inputs

#### U.S. population {-}

Just need to round the population number the nearest whole number. Then export data to FINAL folder.

```{r, results='hide', warning=FALSE, message=FALSE}

pop <- read_xlsx("data_inputs/OTHER/us_population/DATA/NHANES 17 SUBPOPratio_0327.xlsx", sheet = "final")

pop1 <- pop %>% rename(subgroup = subgroup_id,
                       `2018_pop` = Pop_2018) %>% 
  mutate(`2018_pop` = ceiling(`2018_pop`)) # round up to next whole number

write_csv(pop1, paste0("data_inputs/FINAL/cleaned_raw_data/population_distribution", my_date))

```

#### Conversion units {-}

Just select the needed variables then export.

```{r, results='hide', warning=FALSE, message=FALSE}

units <- read_csv("data_inputs/OTHER/unit_conversions/DATA/Unit_conversions_1.4.24.csv")

units1 <- units %>% select(Food_group, DGA_unit, Conversion_to_grams, Equation)

write_csv(units1, paste0("data_inputs/FINAL/cleaned_raw_data/unit_conversions", my_date))

```

Clear global environment expect for my_date.

```{r, results='hide', warning=FALSE, message=FALSE}

rm(list=setdiff(ls(), c("my_date")))

```

### Health Data Inputs

#### Cancer incidence {-}

Import the raw cancer data input, clean up the variable names, and calculate the standard error of the cancer counts ('count_se').

```{r, results='hide', warning=FALSE, message=FALSE}

# import cancer data
cancer <- read_xls("data_inputs/HEALTH/cancer_incidence/DATA/2018CANCERRATE_0327.xls", sheet = "2018pop")

cancer1 <- 
  cancer %>% select(subgroup_id, cancer_code_ICD_O_3, Crude_Rate, 
                  Standard_Error, `_2018_pop`, No_2018) %>% 
  rename(crude_rate = Crude_Rate,
                  crude_se = Standard_Error,
                  count = No_2018,
                  population = `_2018_pop`,
         subgroup = subgroup_id,
         diseases = cancer_code_ICD_O_3) %>% 
  mutate(count_se = (crude_se / 100000) * population)

```

Recode the cancer labels as shorter abbreviations.

```{r, results='hide', warning=FALSE, message=FALSE}

cancer2 <- cancer1 %>% mutate(diseases = recode(diseases, 
                                     "All Sites" = "ALL",
                                     "Colon and Rectum" = "CC",
                                     "Corpus Uteri" = "UC",
                                     "Esophagus" = "ECA",
                                     "Breast" = "BC",
                                     "Gallbladder" = "GC",
                                     "Kidney and Renal Pelvis" = "KC",
                                     "Liver and Intrahepatic Bile Duct" = "LVC",
                                     "Lung and Bronchus" = "LC",
                                     "myeloma" = "MMC",
                                     "Ovary" = "OC",
                                     "Pancreas" = "PC",
                                     "prostate(advanced)" = "APCA",
                                     "stomach_cardia" = "SCC",
                                     "stomach_noncar" = "SCNC",
                                     "Thyroid" = "TC",
                                     "oral cavity and pharynx and larynx" = "MLPC"))

```

Round the count values to the nearest whole number.

```{r, results='hide', warning=FALSE, message=FALSE}

cancer3 <- cancer2 %>% mutate(population = ceiling(population),
                              count = ceiling(count),
                              count_se = ceiling(count_se)) %>% 
  arrange(diseases, subgroup)

```

Lastly, merge with population subsgroup information and cancer labels.

```{r, results='hide', warning=FALSE, message=FALSE}

# import subgroups info
subgrps <- read_csv("data_inputs/OTHER/labels/DATA/population_subgroups_48_060923_FINAL.csv")

# join with cancer data
cancer4 <- left_join(cancer3, subgrps, by = "subgroup") 

# import cancer label data
dis_labels <- read_csv("data_inputs/OTHER/labels/DATA/disease_outcomes_060923_FINAL.csv") %>% 
  select(outcome, outcome_label)

cancer5 <- left_join(cancer4, dis_labels, by = c("diseases" = "outcome")) %>% 
  rename(diseases_label = outcome_label) %>% 
  relocate(diseases_label, .after = diseases)

```

If the rate or standard error variable are missing, then replace with 0. Lastly, export the cleaned file to the "FINAL" folder.

```{r, results='hide', warning=FALSE, message=FALSE}

cancer6 <- cancer5 %>% mutate(crude_rate = ifelse(is.na(crude_rate), 0, crude_rate),
                              crude_se = ifelse(is.na(crude_se), 0, crude_se),
                              count_se = ifelse(is.na(count_se), 0, count_se))

# export final dataset
write_csv(cancer6, paste0("data_inputs/FINAL/cleaned_raw_data/cancer_incidence", my_date))

```

#### Cardiovascular disease (CVD) mortaliaty {-}

Import raw CVD data and tidy.

```{r, results='hide', warning=FALSE, message=FALSE}

cvd <- read_xlsx("data_inputs/HEALTH/cvd_mortality/DATA/CVDmotality2018_11022023.xlsx", sheet = "merge")

cvd1 <- cvd %>% select(cause, subgroup_id, Deaths, Deaths_se) %>% 
  rename(subgroup = subgroup_id) %>% 
  mutate(deaths_rounded = ceiling(Deaths),
         deaths_se_rounded = ceiling(Deaths_se))

# look at outcome labels
unique(cvd1$cause)

# need to rename diabetes
cvd2 <- cvd1 %>% mutate(cause = recode(cause, 
                               "DM" = "DIAB")) %>% 
  select(-c(Deaths, Deaths_se))

```

Transform the mean and SE data from long to wide format and then rejoin.

```{r, results='hide', warning=FALSE, message=FALSE}

# just mean death values
cvd_wide_mean <- pivot_wider(cvd2 %>% select(-deaths_se_rounded), 
                        names_from = cause, 
                        values_from = deaths_rounded)

# just se death values
cvd_wide_se <- pivot_wider(cvd2 %>% select(-deaths_rounded), 
                           names_from = cause, 
                           values_from = deaths_se_rounded,
                           names_prefix = "se_")

# merge
cvd_wide <- left_join(cvd_wide_mean, cvd_wide_se, by = "subgroup")

```

Join with population subgroup information.

```{r, results='hide', warning=FALSE, message=FALSE}

# get rid of NA row
cvd_wide1 <- cvd_wide %>% filter(!(is.na(subgroup)))

# merge with cvd data
cvd_final <- left_join(cvd_wide1, subgrps, by = "subgroup") %>% 
  relocate(subgroup, Age, Age_label, Sex, Sex_label, Race, Race_label)

```

Create the BMI-mediated and SBP-mediated variables by simply setting them equal to the cvd death values. This is needed for the model code to work properly later.

```{r, results='hide', warning=FALSE, message=FALSE}

cvd_final1 <- cvd_final %>% mutate(
                      
                     # BMI
  
                     AA_medBMI = AA,
                     se_AA_medBMI = se_AA,
                                   
                     AFF_medBMI = AFF,
                     se_AFF_medBMI = se_AFF,
                     
                     CM_medBMI = CM,
                     se_CM_medBMI = se_CM,
                     
                     DIAB_medBMI = DIAB,
                     se_DIAB_medBMI = se_DIAB,
                     
                     ENDO_medBMI = ENDO,
                     se_ENDO_medBMI = se_ENDO,
                     
                     HHD_medBMI = HHD,
                     se_HHD_medBMI = se_HHD,
                     
                     HSTK_medBMI = HSTK,
                     se_HSTK_medBMI = se_HSTK,
                     
                     IHD_medBMI = IHD,
                     se_IHD_medBMI = se_IHD,
                     
                     ISTK_medBMI = ISTK,
                     se_ISTK_medBMI = se_ISTK,
                     
                     OSTK_medBMI = OSTK,
                     se_OSTK_medBMI = se_OSTK,
                     
                     OTH_medBMI = OTH,
                     se_OTH_medBMI = se_OTH,
                     
                     PVD_medBMI = PVD,
                     se_PVD_medBMI = se_PVD,
                     
                     RHD_medBMI = RHD,
                     se_RHD_medBMI = se_RHD,
                     
                     TSTK_medBMI = TSTK,
                     se_TSTK_medBMI = se_TSTK,
                     
                     # SBP
                     
                     AA_medSBP = AA,
                     se_AA_medSBP = se_AA,
                     
                     AFF_medSBP = AFF,
                     se_AFF_medSBP = se_AFF,
                     
                     CM_medSBP = CM,
                     se_CM_medSBP = se_CM,
                     
                     DIAB_medSBP = DIAB,
                     se_DIAB_medSBP = se_DIAB,
                     
                     ENDO_medSBP = ENDO,
                     se_ENDO_medSBP = se_ENDO,
                     
                     HHD_medSBP = HHD,
                     se_HHD_medSBP = se_HHD,
                     
                     HSTK_medSBP = HSTK,
                     se_HSTK_medSBP = se_HSTK,
                     
                     IHD_medSBP = IHD,
                     se_IHD_medSBP = se_IHD,
                     
                     ISTK_medSBP = ISTK,
                     se_ISTK_medSBP = se_ISTK,
                     
                     OSTK_medSBP = OSTK,
                     se_OSTK_medSBP = se_OSTK,
                     
                     OTH_medSBP = OTH,
                     se_OTH_medSBP = se_OTH,
                     
                     PVD_medSBP = PVD,
                     se_PVD_medSBP = se_PVD,
                     
                     RHD_medSBP = RHD,
                     se_RHD_medSBP = se_RHD,
                     
                     TSTK_medSBP = TSTK,
                     se_TSTK_medSBP = se_TSTK)

```

Export to FINAL folder.

```{r, results='hide', warning=FALSE, message=FALSE}

write_csv(cvd_final1, paste0("data_inputs/FINAL/cleaned_raw_data/cvd_mortality", my_date))

```

#### Other health data {-}

Import other health datsets.

```{r, results='hide', warning=FALSE, message=FALSE}

weight <- read_xls("data_inputs/HEALTH/overweight_prevalence/DATA/overweight1518_48grp_04.10.23.xls")

hbp <- read_xls("data_inputs/HEALTH/systolic_blood_pressure/DATA/SBP_1718_04192023.xls", sheet = "HBPbyGRP")

sbp <- read_xls("data_inputs/HEALTH/systolic_blood_pressure/DATA/SBP_1718_04192023.xls", sheet = "over20yr")

#high sbp variables
highSBP <- read_xlsx("data_inputs/HEALTH/systolic_blood_pressure/DATA/highSBP_data.xlsx")

```

Create non-Hipsanic Black (NHB) variables.

```{r, results='hide', warning=FALSE, message=FALSE}

oth_health <- weight %>% mutate(nhb = ifelse(Race_label == "NHB", 1, 0),
                            nhb_se = 0) %>% 
  rename(overweight_rate = Percent,
         overweight_rate_se = StdErr) %>% 
  select(subgroup, overweight_rate, overweight_rate_se, nhb, nhb_se)

```

The SBP data is at the participant-level so it needs to be summarized at population subgroup level.

```{r, results='hide', warning=FALSE, message=FALSE}

sbp1 <- sbp %>% 
  filter(!(is.na(mean_sbp))) # only include those with non-missing mean_sbp value

# summarize by subgroup ID
sbp_vars <- sbp1 %>% 
  group_by(subgroup_id) %>% 
  summarise(sbp_mean = mean(mean_sbp),
            # calculate standard error
            sbp_se = sd(mean_sbp)/sqrt(length(mean_sbp)))

# merge with highSBP
highSBP1 <- highSBP %>% 
  select(subgroup_id, Percent, StdErr) %>% 
  rename(highSBP_rate = Percent,
         highSBP_rate_se = StdErr)

sbp_vars1 <- left_join(sbp_vars, highSBP1, by = "subgroup_id")

# merge sbp vars with oth_health
oth_health1 <- left_join(oth_health, sbp_vars1, by = c("subgroup" = "subgroup_id"))

# hbp variables
hbp1 <- 
  hbp %>% rename(hbp = Percent,
               hbp_se = StdErr) %>% 
  mutate(hbp = hbp/100,
         hbp_se = hbp_se/100) %>% 
  select(subgroup_id, hbp, hbp_se)

# merge hbp vars with oth_health
oth_health2 <- left_join(oth_health1, hbp1, by = c("subgroup" = "subgroup_id"))

```

Export to FINAL folder.

```{r, results='hide', warning=FALSE, message=FALSE}

write_csv(oth_health2, paste0("data_inputs/FINAL/cleaned_raw_data/other_health", my_date))

```

#### Effect sizes for diet and body mass index (BMI) {-}

No changes are needed so just import the file then export to FINAL folder.

```{r, results='hide', warning=FALSE, message=FALSE}

bmi_effects <- read_csv("data_inputs/HEALTH/effect_sizes_dietfactor_bmi/DATA/food_to_BMI_effects_from_Dari_1.5.24.csv")

write_csv(bmi_effects, paste0("data_inputs/FINAL/cleaned_raw_data/food_to_bmi_effects", my_date))

```

#### Effect sizes for diet and systolic blood pressure (SBP) {-}

No changes are needed so just import the file then export to FINAL folder.

```{r, results='hide', warning=FALSE, message=FALSE}

sbp_effects <- read_csv("data_inputs/HEALTH/effect_sizes_dietfactor_sbp/DATA/food_to_sbp_effects_NOT_converted_1.5.24.csv")

write_csv(sbp_effects, paste0("data_inputs/FINAL/cleaned_raw_data/food_to_sbp_effects", my_date))

```

#### Relative risks (RR) for BMI and cancer {-}

No changes are needed so just import the file then export to FINAL folder.

```{r, results='hide', warning=FALSE, message=FALSE}

bmi_cancer <- read_csv("data_inputs/HEALTH/rr_bmi_cancer/DATA/RR_BMI_cancer_11.23.22.csv") 

write_csv(bmi_cancer, paste0("data_inputs/FINAL/cleaned_raw_data/rr_bmi_cancer", my_date))

```

#### Relative risks (RR) for BMI and CVD {-}

No changes are needed so just import the file then export to FINAL folder.

```{r, results='hide', warning=FALSE, message=FALSE}

bmi_cvd <- read_csv("data_inputs/HEALTH/rr_bmi_cvd/DATA/RR_BMI_cvd_3.2.23.csv")

write_csv(bmi_cvd, paste0("data_inputs/FINAL/cleaned_raw_data/rr_bmi_cvd", my_date))

```

#### Relative risks (RR) for SBP and CVD {-}

No changes are needed so just import the file then export to FINAL folder.

```{r, results='hide', warning=FALSE, message=FALSE}

sbp_cvd <- read_csv("data_inputs/HEALTH/rr_sbp_cvd/DATA/RR_sbp_cvd_2.16.23.csv")

write_csv(sbp_cvd, paste0("data_inputs/FINAL/cleaned_raw_data/rr_sbp_cvd", my_date))

```

#### Log relative risks (LogRR) for diet-CVD {-}

No changes are needed so just import the file then export to FINAL folder.

```{r, results='hide', warning=FALSE, message=FALSE}

rm(list=ls(pattern="^cvd"))

cvd <- read_csv("data_inputs/HEALTH/logrr_dietfactor_disease/DATA/logRR_diet_cvd_byage_1.26.24.csv")

write_csv(cvd, paste0("data_inputs/FINAL/cleaned_raw_data/logRR_diet_cvd", my_date))

```

#### Log relative risks (LogRR) for diet-cancer {-}

No changes are needed so just import the file then export to FINAL folder.

```{r, results='hide', warning=FALSE, message=FALSE}

rm(list=ls(pattern="^cancer"))

cancer <- read_csv("data_inputs/HEALTH/logrr_dietfactor_disease/DATA/logRR_diet_cancer_1.5.24.csv")

write_csv(cancer, paste0("data_inputs/FINAL/cleaned_raw_data/logRR_diet_cancer", my_date))

```

#### TMRED {-}

Import the TMRED values (in grams). Then, import the conversion units and join with the TMRED data.

```{r, results='hide', warning=FALSE, message=FALSE}

tmred_g <- read_csv("data_inputs/HEALTH/tmred/DATA/TMRED_grams_1.5.24.csv")

# import conversion units
conversion <- read_csv(paste0("data_inputs/FINAL/cleaned_raw_data/unit_conversions", my_date))

# join with tmred dataset
tmred_dga <- left_join(tmred_g, conversion, by = c("Risk_factor" = "Food_group"))

```

Then, transform the TMRED mean and standard deviation values from grams to "FPED units" (i.e., cups or ounces).

```{r, results='hide', warning=FALSE, message=FALSE}

tmred_dga1 <- tmred_dga %>% mutate(tmred_dga_units = TMRED / Conversion_to_grams,
                     sd_dga_units = SD / Conversion_to_grams)

tmred_dga2 <- tmred_dga1 %>% 
  select(Risk_factor, tmred_dga_units, sd_dga_units, DGA_unit) %>% 
  rename(TMRED = tmred_dga_units,
         SD = sd_dga_units,
         Unit = DGA_unit) %>% 
  mutate(TMRED = signif(TMRED, 3), # round to 3 sig figs
         SD = signif(SD, 3), # round to 3 sig figs
    Unit = str_sub(Unit, start = 3)) # fix unit var

```

Then, export the conversion units in both grams and FPED units to be used later.

```{r, results='hide', warning=FALSE, message=FALSE}

# export grams dataset
write_csv(tmred_g, paste0("data_inputs/FINAL/cleaned_raw_data/tmred_grams", my_date))

# export FPED units
write_csv(tmred_dga2, paste0("data_inputs/FINAL/cleaned_raw_data/tmred_dga_units", my_date))

```

Clear global environment expect for my_date.

```{r, results='hide', warning=FALSE, message=FALSE}

rm(list=setdiff(ls(), c("my_date")))

```

### Diet Data Inputs

#### Dietary intake {-}

Import the NHANES diet intake dataset and clean up.

```{r, results='hide', warning=FALSE, message=FALSE}

nhanes <- read_csv("data_inputs/DIET/dietary_intake/DATA/output_data_from_cluster/NHANES_1518_summary_allfoods_adj_bysub_ncimethod_07-08-2024.csv")

nhanes1 <- nhanes %>% 
  rename("Foodgroup" = "food",
         "Mean_Intake" = "mean",
         "SE_Intake" = "SE",
         "Food_label" = "food_label",
         "Food_desc" = "food_desc") %>% 
  mutate(Food_label = ifelse(Foodgroup == "kcal", "Kilocalorires", Food_label),
         Food_desc = ifelse(Foodgroup == "kcal", "Energy (kcal)", Food_desc)) %>% 
  select(-c(starts_with(c("gro_", "oth_"))))

```

Then, convert the sugar sweetened beverage (SSB) mean and standard error from grams to cup (8 fl oz).

```{r, results='hide', warning=FALSE, message=FALSE}

nhanes2 <- nhanes1 %>% 
  # Divide by 240 to go from grams to cup
  mutate(Mean_Intake = ifelse(Foodgroup == "ssb", Mean_Intake/240, Mean_Intake),
  SE_Intake = ifelse(Foodgroup == "ssb", SE_Intake/240, SE_Intake),
  Food_desc = ifelse(Foodgroup == "ssb", "Sugar sweetened beverages (1 cup [8 fl oz])", Food_desc))

nhanes3 <- nhanes2 %>% arrange(Foodgroup, subgroup)

```

Lastly, rename the standard deviation variable and then export.

```{r, results='hide', warning=FALSE, message=FALSE}

# rename stddev
nhanes4 <- nhanes3 %>% 
  rename(sigma_u_wgt = StdDev) %>%
  mutate(sigma_u_wgt = ifelse(Foodgroup == "ssb", sigma_u_wgt/240, sigma_u_wgt)) %>%
  relocate(sigma_u_wgt, .after = "SE_Intake")

# export
write_csv(nhanes4, paste0("data_inputs/FINAL/cleaned_raw_data/nhanes1518_agesexrace", my_date))

```

#### Counterfactual intake {-}

No changes are needed so just import the file then export to FINAL folder.

```{r, results='hide', warning=FALSE, message=FALSE}

cf <- read_csv("data_inputs/DIET/counterfactual_intake/DATA/counterfactual_intake_050724.csv")

write_csv(cf, paste0("data_inputs/FINAL/cleaned_raw_data/counterfactual_intake", my_date))

```

