# Calculate Cost, Environment, and Social Impact Factors

This chapter walks you through all of the R code used to calculate the impact factors that are used in the final model. Additionally, the food group-level inedible and wasted proportions are calculated at the end.

Note that you must first open the 'methods_manual' R project before running this script or else it will not work.

First, let's set up our environment.

```{r}

# check working directory
getwd()

# SET UP -----

rm(list = ls())

options(scipen=999)

library(tidyverse)
library(readxl)
library(survey)
library(bookdown)

my_date <- Sys.Date()

```

## Prepare Data

First, we need to import the NHANES dataset to get the survey weight variables that are necessary to incorporate when calculating mean estimates from NHANES data. See more information [here](https://wwwn.cdc.gov/nchs/nhanes/tutorials/weighting.aspx).

```{r}

# import subgroup-seqn mapping
nhanes <- read_rds("data_inputs/DIET/dietary_intake/DATA/clean_data/nhanes1518_adj_clean_wide.rds")

# select subset of variables
subgroup_dat <- nhanes %>% select(SEQN, subgroup, SDMVPSU, SDMVSTRA, wtnew, inAnalysis)

```

Then, create a cleaning function that will be used to clean the output summary data from the "svymean" function used below. 

```{r}

clean_func <- function(x, y){
  
  x1 <- x %>% select(subgroup, contains(y))
  
  # split up into mean and se
  my_se <- x1 %>% select(subgroup, starts_with("se."))
  my_mean <- x1 %>% select(subgroup, !starts_with("se."))
  
  # transform both datsets to long
  my_se_long <- my_se %>% pivot_longer(cols = starts_with("se."),
                                       names_to = "food",
                                       names_prefix = c("se."),
                                       values_to = paste0(y, "_se"))
  
  my_mean_long <- my_mean %>% pivot_longer(cols = !subgroup,
                                           names_to = "food",
                                           values_to = paste0(y, "_mean"))
  
  # join
  allfoods_bysub_long <- left_join(my_mean_long, my_se_long, by = c("subgroup", "food"))
  
  # fix names
  dat <- allfoods_bysub_long %>% 
    mutate(food_type = case_when(grepl("fah", food) ~ "Grocery",
                                 grepl("fafh", food) ~ "Non-Grocery",
                                 grepl("total", food) ~ "Total")) %>% 
    mutate(food = gsub(".*_sum_", "", food)) %>% 
    arrange(subgroup, food)
  
  print(dat)
  
}

```

Import the data that were cleaned in the last few chapters.

```{r}

# diet intake data
# both_days <- read_rds("data_inputs/IMPACT_FACTORS/temp_data/both_days15_env.rds")

# price data
price_dat <- read_rds("data_inputs/IMPACT_FACTORS/temp_data/price_input_dat.rds")

# environment and social data
enviro_dat <- read_rds("data_inputs/IMPACT_FACTORS/temp_data/enviro_input_dat.rds")

```

## Cost Impact Factors {#cost-impact-factors}

### Cost Impact Factors, by Food, by Subgroup

Join the price data with the NHANES subgroup data.

```{r}

# full join
price_dat1 <- full_join(price_dat, subgroup_dat, by = c("seqn" = "SEQN"))

```

Define the "survey design" for the cost dataset. Survey designs are a unique part of the R 'survey' package. Set the appropriate weight variables.

Then, subset the dataset to only include participants who are eligible to be in the sample (i.e., inAnalysis == 1). Chapter \@ref(step-seven) explains more about these weight and inAnalysis variables.

```{r}

# Define survey design for cost dataset 
my_cost_svy <- svydesign(data=price_dat1, 
                    id=~SDMVPSU, # Masked Variance Unit Pseudo-PSU 
                    strata=~SDMVSTRA, # Masked Variance Unit Pseudo-Stratum 
                    weights=~wtnew, # New sample weight
                    nest=TRUE)

# Create a survey design object for the subset of interest 
my_cost_svy_sub <- subset(my_cost_svy, inAnalysis==1)

```

#### Calculate Daily Mean Cost, For The Whole Sample

Apply the 'svymean' function in order to calculate the mean daily cost for each food group, using the entire sample. 

```{r}

allfoods_cost <- svymean(reformulate(names(price_dat1) %>% str_subset("price")),
                             my_cost_svy_sub) %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "var") %>% 
  pivot_wider(names_from = var,
              values_from = c(mean, SE)) %>% 
  mutate(subgroup = 0) %>% 
  relocate(subgroup)

# fix colnames
colnames(allfoods_cost) <- gsub("mean_", "", colnames(allfoods_cost))
colnames(allfoods_cost) <- gsub("SE_", "se.", colnames(allfoods_cost))

```

#### Calculate Daily Mean Consumed, Inedible, and Wasted Amounts (FNDDS-Level), For The Whole Sample

Apply the 'svymean' function in order to calculate the mean daily consumed, inedible, and wasted amounts of each food group, using the entire sample. 

##### Consumed

```{r}

allfoods_cost_consumed <- svymean(reformulate(names(price_dat1) %>% str_subset("consumed")),
                                  my_cost_svy_sub) %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "var") %>% 
  pivot_wider(names_from = var,
              values_from = c(mean, SE)) %>% 
  mutate(subgroup = 0) %>% 
  relocate(subgroup)

# fix colnames
colnames(allfoods_cost_consumed) <- gsub("mean_", "", colnames(allfoods_cost_consumed))
colnames(allfoods_cost_consumed) <- gsub("SE_", "se.", colnames(allfoods_cost_consumed))

```

##### Inedible

```{r}

allfoods_cost_inedible <- svymean(reformulate(names(price_dat1) %>% str_subset("inedible")),
                                  my_cost_svy_sub) %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "var") %>% 
  pivot_wider(names_from = var,
              values_from = c(mean, SE)) %>% 
  mutate(subgroup = 0) %>% 
  relocate(subgroup)

# fix colnames
colnames(allfoods_cost_inedible) <- gsub("mean_", "", colnames(allfoods_cost_inedible))
colnames(allfoods_cost_inedible) <- gsub("SE_", "se.", colnames(allfoods_cost_inedible))

```

##### Wasted

```{r}

allfoods_cost_wasted <- svymean(reformulate(names(price_dat1) %>% str_subset("wasted")),
                                my_cost_svy_sub) %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "var") %>% 
  pivot_wider(names_from = var,
              values_from = c(mean, SE)) %>% 
  mutate(subgroup = 0) %>% 
  relocate(subgroup)

# fix colnames
colnames(allfoods_cost_wasted) <- gsub("mean_", "", colnames(allfoods_cost_wasted))
colnames(allfoods_cost_wasted) <- gsub("SE_", "se.", colnames(allfoods_cost_wasted))

```

#### Calculate Daily Mean Cost, For All 48 Subgroups

Apply the 'svyby' function in order to calculate the mean daily cost for each food group, separately for each of the 48 population subgroups.

```{r}

allfoods_cost_bysub <- svyby(reformulate(names(price_dat1) %>% str_subset("price")),
                            ~subgroup,
                            my_cost_svy_sub,
                            svymean)

```

Combine the mean cost values for the whole sample (i.e., subgroup 0) with the mean cost values for subgroups 1-48.

```{r}

# bind with subgroup 0
allfoods_cost_bysub1 <- rbind(allfoods_cost, allfoods_cost_bysub)

```

#### Calculate Daily Mean Consumed, Inedible, and Wasted Amounts (FNDDS-Level), For All 48 Subgroups

Apply the 'svyby' function in order to calculate the mean daily consumed, inedible, and wasted amounts of each food group, separately for each of the 48 population subgroups.

Then, combine the mean values for the whole sample (i.e., subgroup 0) with the mean values for subgroups 1-48.

##### Consumed

```{r}

allfoods_cost_consumed_bysub <- svyby(reformulate(names(price_dat1) %>% str_subset("consumed")),
                             ~subgroup,
                             my_cost_svy_sub,
                             svymean)

# bind with subgroup 0
allfoods_cost_consumed_bysub1 <- rbind(allfoods_cost_consumed, allfoods_cost_consumed_bysub)

```

##### Inedible

```{r}

allfoods_cost_inedible_bysub <- svyby(reformulate(names(price_dat1) %>% str_subset("inedible")),
                                    ~subgroup,
                                    my_cost_svy_sub,
                                    svymean)

# bind with subgroup 0
allfoods_cost_inedible_bysub1 <- rbind(allfoods_cost_inedible, allfoods_cost_inedible_bysub)

```

##### Wasted

```{r}

allfoods_cost_wasted_bysub <- svyby(reformulate(names(price_dat1) %>% str_subset("wasted")),
                                      ~subgroup,
                                      my_cost_svy_sub,
                                      svymean)

# bind with subgroup 0
allfoods_cost_wasted_bysub1 <- rbind(allfoods_cost_wasted, allfoods_cost_wasted_bysub)

```

#### Clean Output Data

Apply the cleaning function (defined above) to the mean cost dataset.

```{r}

# APPLY FUNCTION TO COST DAT
cost_dat <- clean_func(x = allfoods_cost_bysub1, y = "price")

# MORE MANUAL CLEANING
cost_dat1 <- 
  cost_dat %>% mutate(intake_type = case_when(grepl("Consumed", food) ~ "Consumed",
                                            grepl("Wasted", food) ~ "Wasted",
                                            grepl("Inedible", food) ~ "Inedible")) %>% 
  mutate(food = gsub("Consumed_|Wasted_|Inedible_", "", food))

# MORE MANUAL CLEANING
cost_dat2 <- 
  cost_dat1 %>% mutate(mixed_dish = case_when(grepl("_0", food) ~ "Non-Mixed",
                                              grepl("_1", food) ~ "Mixed")) %>% 
  mutate(food = gsub("_0|_1", "", food)) %>% 
  relocate(intake_type, food_type, mixed_dish, .after = food) %>% 
  arrange(intake_type, subgroup, food, food_type, mixed_dish)

# examine
cost_dat2 %>% filter(food_type == "Total") %>% head()
  
# pivot to wide
cost_dat3 <-
  cost_dat2 %>% pivot_wider(names_from = intake_type,
                          values_from = c(price_mean, price_se)) %>% 
  arrange(subgroup, food, food_type)

```

Apply the cleaning function (defined above) to the mean consumed, inedible, and wasted datasets.

```{r}

# APPLY FUNCTION TO INTAKE DATASETS
cost_consumed_dat <- clean_func(x = allfoods_cost_consumed_bysub1, y = "consumed")
cost_wasted_dat <- clean_func(x = allfoods_cost_wasted_bysub1, y = "wasted")
cost_inedible_dat <- clean_func(x = allfoods_cost_inedible_bysub1, y = "inedible")

# join all 3
cost_intake_dat <- left_join(cost_consumed_dat, cost_wasted_dat, by = c("subgroup", "food", "food_type")) %>% 
  left_join(cost_inedible_dat, by = c("subgroup", "food", "food_type"))

# create mixed dish variable
cost_intake_dat1 <- cost_intake_dat %>% 
  mutate(mixed_dish = case_when(grepl("_0", food) ~ "Non-Mixed",
                                grepl("_1", food) ~ "Mixed")) %>% 
  mutate(food = gsub("_0|_1", "", food))

# join
cost_final_dat <- 
  left_join(cost_dat3, cost_intake_dat1, by = c("subgroup", "food", "food_type", "mixed_dish")) %>% 
  rename(fndds_consumed_mean = consumed_mean,
         fndds_consumed_se = consumed_se,
         fndds_wasted_mean = wasted_mean,
         fndds_wasted_se = wasted_se,
         fndds_inedible_mean = inedible_mean,
         fndds_inedible_se = inedible_se)

```

### Cost Impact Factors, Per 100 Grams and Per 1 DGA Serving

First, import the conversion units.

```{r}

units <- read_csv(paste("data_inputs/FINAL/cleaned_raw_data/unit_conversions_", my_date, "_FINAL.csv", sep="")) %>% 
  select(Food_group, Conversion_to_grams)

# merge with units
cost_final_dat1 <- cost_final_dat %>% 
  left_join(units, by = c("food" = "Food_group"))

cost_final_dat1 %>% filter(is.na(Conversion_to_grams)) %>% head()

```

Then, calculate the cost impact factors (i) per 100 grams and (ii) per DGA unit.

To calculate the cost per 100 grams, you divide the daily mean cost of the food group by the daily amount of food consumed, and then multiply by 100.

```{r}

# calculate impact factors
cost_final_dat2 <- cost_final_dat1 %>% 
  rowwise() %>% 
  mutate(costper100gram_consumed = ifelse(fndds_consumed_mean == 0, 0, (price_mean_Consumed / fndds_consumed_mean) * 100),
         costperDGA_consumed = costper100gram_consumed * (Conversion_to_grams / 100)) 

# check
cost_final_dat2 %>% select(subgroup, food, food_type, mixed_dish, costperDGA_consumed) %>% head()

```

## Environmental & Social Impact Factors {#environmental-social-impact-factors}

### Environmental & Social Impact Factors, by Food, by Subgroup

Essentially the same steps that were used to calculate the cost impact factors are also used to calculate the environmental and social impact factors.

First, define the "survey design" for the enviro dataset. Survey designs are a unique part of the R 'survey' package. Set the appropriate weight variables.

Then, subset the dataset to only include participants who are eligible to be in the sample (i.e., inAnalysis == 1). Chapter \@ref(step-seven) explains more about these weight and inAnalysis variables.

```{r}

# join
enviro_dat1 <- full_join(enviro_dat, subgroup_dat, by = c("seqn" = "SEQN"))

# Define survey design for ghg dataset 
my_enviro_svy <- svydesign(data=enviro_dat1, 
                           id=~SDMVPSU, # Masked Variance Unit Pseudo-PSU 
                           strata=~SDMVSTRA, # Masked Variance Unit Pseudo-Stratum 
                           weights=~wtnew, # New sample weight
                           nest=TRUE)

# Create a survey design object for the subset of interest 
my_enviro_svy_sub <- subset(my_enviro_svy, inAnalysis==1)

```

#### Calculate Daily Mean Environment and Social Impacts, (i) For The Whole Sample, and (ii) For All 48 Subgroups

##### Greenhouse Gas Emissions (GHG)

```{r}

# CALCULATE GHG IMPACT (WHOLE SAMPLE)
allfoods_ghg <- svymean(reformulate(names(enviro_dat1) %>% str_subset("ghg")),
                         my_enviro_svy_sub) %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "var") %>% 
  pivot_wider(names_from = var,
              values_from = c(mean, SE)) %>% 
  mutate(subgroup = 0) %>% 
  relocate(subgroup)

# fix colnames
colnames(allfoods_ghg) <- gsub("mean_", "", colnames(allfoods_ghg))
colnames(allfoods_ghg) <- gsub("SE_", "se.", colnames(allfoods_ghg))

# CALCULATE GHG IMPACT, BY SUBGROUP
allfoods_ghg_bysub <- svyby(reformulate(names(enviro_dat1) %>% str_subset("ghg")),
                            ~subgroup,
                            my_enviro_svy_sub,
                            svymean)

# bind with subgroup 0
allfoods_ghg_bysub1 <- rbind(allfoods_ghg, allfoods_ghg_bysub)
```

##### Cumulative Energy Demand (CED)

```{r}

# CALCULATE CED IMPACT (WHOLE SAMPLE)
allfoods_ced <- svymean(reformulate(names(enviro_dat1) %>% str_subset("ced")),
                        my_enviro_svy_sub) %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "var") %>% 
  pivot_wider(names_from = var,
              values_from = c(mean, SE)) %>% 
  mutate(subgroup = 0) %>% 
  relocate(subgroup)

# fix colnames
colnames(allfoods_ced) <- gsub("mean_", "", colnames(allfoods_ced))
colnames(allfoods_ced) <- gsub("SE_", "se.", colnames(allfoods_ced))

# CALCULATE CED IMPACT, BY SUBGROUP
allfoods_ced_bysub <- svyby(reformulate(names(enviro_dat1) %>% str_subset("ced")),
                            ~subgroup,
                            my_enviro_svy_sub,
                            svymean)

# bind with subgroup 0
allfoods_ced_bysub1 <- rbind(allfoods_ced, allfoods_ced_bysub)

```

##### Water Scarcity

```{r}

# CALCULATE WATER IMPACT (WHOLE SAMPLE)
allfoods_water <- svymean(reformulate(names(enviro_dat1) %>% str_subset("^water")),
                        my_enviro_svy_sub) %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "var") %>% 
  pivot_wider(names_from = var,
              values_from = c(mean, SE)) %>% 
  mutate(subgroup = 0) %>% 
  relocate(subgroup)

# fix colnames
colnames(allfoods_water) <- gsub("mean_", "", colnames(allfoods_water))
colnames(allfoods_water) <- gsub("SE_", "se.", colnames(allfoods_water))

# CALCULATE WATER IMPACT, BY SUBGROUP
allfoods_water_bysub <- svyby(reformulate(names(enviro_dat1) %>% str_subset("^water")),
                            ~subgroup,
                            my_enviro_svy_sub,
                            svymean)

# bind with subgroup 0
allfoods_water_bysub1 <- rbind(allfoods_water, allfoods_water_bysub)

```

##### Bluewater Use

```{r}

# CALCULATE BLUEWATER IMPACT (WHOLE SAMPLE)
allfoods_bluewater <- svymean(reformulate(names(enviro_dat1) %>% str_subset("bluewater")),
                          my_enviro_svy_sub) %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "var") %>% 
  pivot_wider(names_from = var,
              values_from = c(mean, SE)) %>% 
  mutate(subgroup = 0) %>% 
  relocate(subgroup)

# fix colnames
colnames(allfoods_bluewater) <- gsub("mean_", "", colnames(allfoods_bluewater))
colnames(allfoods_bluewater) <- gsub("SE_", "se.", colnames(allfoods_bluewater))

# CALCULATE BLUEWATER IMPACT, BY SUBGROUP
allfoods_bluewater_bysub <- svyby(reformulate(names(enviro_dat1) %>% str_subset("bluewater")),
                              ~subgroup,
                              my_enviro_svy_sub,
                              svymean)

# bind with subgroup 0
allfoods_bluewater_bysub1 <- rbind(allfoods_bluewater, allfoods_bluewater_bysub)

```

##### Forced Labor Risk (FL)

```{r}

# CALCULATE FL IMPACT (WHOLE SAMPLE)
allfoods_fl <- svymean(reformulate(names(enviro_dat1) %>% str_subset("fl")),
                        my_enviro_svy_sub) %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "var") %>% 
  pivot_wider(names_from = var,
              values_from = c(mean, SE)) %>% 
  mutate(subgroup = 0) %>% 
  relocate(subgroup)

# fix colnames
colnames(allfoods_fl) <- gsub("mean_", "", colnames(allfoods_fl))
colnames(allfoods_fl) <- gsub("SE_", "se.", colnames(allfoods_fl))

# CALCULATE FL IMPACT, BY SUBGROUP
allfoods_fl_bysub <- svyby(reformulate(names(enviro_dat1) %>% str_subset("fl")),
                            ~subgroup,
                            my_enviro_svy_sub,
                            svymean)

# bind with subgroup 0
allfoods_fl_bysub1 <- rbind(allfoods_fl, allfoods_fl_bysub)

```

#### Calculate Daily Mean Consumed, Inedible, and Wasted Amounts (FCID-Level), For The Whole Sample

Apply the 'svymean' function in order to calculate the mean daily consumed, inedible, and wasted amounts of each food group, using the entire sample. 

##### Consumed

```{r}
# consumed
allfoods_enviro_consumed <- svymean(reformulate(names(enviro_dat1) %>% str_subset("consumed")),
                                  my_enviro_svy_sub) %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "var") %>% 
  pivot_wider(names_from = var,
              values_from = c(mean, SE)) %>% 
  mutate(subgroup = 0) %>% 
  relocate(subgroup)

# fix colnames
colnames(allfoods_enviro_consumed) <- gsub("mean_", "", colnames(allfoods_enviro_consumed))
colnames(allfoods_enviro_consumed) <- gsub("SE_", "se.", colnames(allfoods_enviro_consumed))
```

##### Inedible

```{r}
# inedible
allfoods_enviro_inedible <- svymean(reformulate(names(enviro_dat1) %>% str_subset("inedible")),
                                  my_enviro_svy_sub) %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "var") %>% 
  pivot_wider(names_from = var,
              values_from = c(mean, SE)) %>% 
  mutate(subgroup = 0) %>% 
  relocate(subgroup)

# fix colnames
colnames(allfoods_enviro_inedible) <- gsub("mean_", "", colnames(allfoods_enviro_inedible))
colnames(allfoods_enviro_inedible) <- gsub("SE_", "se.", colnames(allfoods_enviro_inedible))
```

##### Wasted

```{r}
# wasted
allfoods_enviro_wasted <- svymean(reformulate(names(enviro_dat1) %>% str_subset("wasted")),
                                    my_enviro_svy_sub) %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "var") %>% 
  pivot_wider(names_from = var,
              values_from = c(mean, SE)) %>% 
  mutate(subgroup = 0) %>% 
  relocate(subgroup)

# fix colnames
colnames(allfoods_enviro_wasted) <- gsub("mean_", "", colnames(allfoods_enviro_wasted))
colnames(allfoods_enviro_wasted) <- gsub("SE_", "se.", colnames(allfoods_enviro_wasted))
```

#### Calculate Daily Mean Consumed, Inedible, and Wasted Amounts (FCID-Level), For All 48 Subgroups

Apply the 'svyby' function in order to calculate the mean daily consumed, inedible, and wasted amounts of each food group, separately for each of the 48 population subgroups.

Then, combine the mean values for the whole sample (i.e., subgroup 0) with the mean values for subgroups 1-48.

##### Consumed

```{r}

allfoods_enviro_consumed_bysub <- svyby(reformulate(names(enviro_dat1) %>% str_subset("consumed")),
                                    ~subgroup,
                                    my_enviro_svy_sub,
                                    svymean)

# bind with subgroup 0
allfoods_enviro_consumed_bysub1 <- rbind(allfoods_enviro_consumed, allfoods_enviro_consumed_bysub)

```

##### Inedible

```{r}
# inedible
allfoods_enviro_inedible_bysub <- svyby(reformulate(names(enviro_dat1) %>% str_subset("inedible")),
                                      ~subgroup,
                                      my_enviro_svy_sub,
                                      svymean)

# bind with subgroup 0
allfoods_enviro_inedible_bysub1 <- rbind(allfoods_enviro_inedible, allfoods_enviro_inedible_bysub)
```

##### Wasted

```{r}
# wasted
allfoods_enviro_wasted_bysub <- svyby(reformulate(names(enviro_dat1) %>% str_subset("wasted")),
                                        ~subgroup,
                                        my_enviro_svy_sub,
                                        svymean)

# bind with subgroup 0
allfoods_enviro_wasted_bysub1 <- rbind(allfoods_enviro_wasted, allfoods_enviro_wasted_bysub)
```

#### Clean Output Data

Apply the cleaning function (defined above) to the mean enviro datasets.

```{r}

# APPLY TO ENVIRO DATA
ghg_dat <- clean_func(x = allfoods_ghg_bysub1, y = "ghg") %>% select(-food_type)
ced_dat <- clean_func(x = allfoods_ced_bysub1, y = "ced") %>% select(-food_type)
water_dat <- clean_func(x = allfoods_water_bysub1, y = "water") %>% select(-food_type)
bluewater_dat <- clean_func(x = allfoods_bluewater_bysub1, y = "bluewater") %>% select(-food_type)
fl_dat <- clean_func(x = allfoods_fl_bysub1, y = "fl") %>% select(-food_type)

# join cleaned datasets
enviro_dat <- 
  left_join(ghg_dat, ced_dat, by = c("subgroup", "food")) %>% 
  left_join(water_dat, by = c("subgroup", "food")) %>% 
  left_join(bluewater_dat, by = c("subgroup", "food")) %>% 
  left_join(fl_dat, by = c("subgroup", "food"))

# more cleaning
enviro_dat1 <- 
  enviro_dat %>% mutate(intake_type = case_when(grepl("Consumed", food) ~ "Consumed",
                                              grepl("Wasted", food) ~ "Wasted",
                                              grepl("Inedible", food) ~ "Inedible")) %>% 
  mutate(food = gsub("Consumed_|Wasted_|Inedible_", "", food))

# pivot to wide
enviro_dat2 <-
  enviro_dat1 %>% pivot_wider(names_from = intake_type,
                            values_from = !c(subgroup, food, intake_type)) %>% 
  arrange(subgroup, food)

```

Apply the cleaning function (defined above) to the mean consumed, inedible, and wasted datasets.

```{r}

# APPLY FUNCTION TO INTAKE DAT
enviro_consumed_dat <- clean_func(x = allfoods_enviro_consumed_bysub1, y = "consumed") %>% select(-food_type)
enviro_wasted_dat <- clean_func(x = allfoods_enviro_wasted_bysub1, y = "wasted") %>% select(-food_type)
enviro_inedible_dat <- clean_func(x = allfoods_enviro_inedible_bysub1, y = "inedible") %>% select(-food_type)

# join
enviro_intake_dat <- left_join(enviro_consumed_dat, enviro_wasted_dat, by = c("subgroup", "food")) %>% 
  left_join(enviro_inedible_dat, by = c("subgroup", "food"))

# join enviro and intake dat
enviro_final_dat <- 
  left_join(enviro_dat2, enviro_intake_dat, by = c("subgroup", "food")) %>% 
  rename(fcid_consumed_mean = consumed_mean,
         fcid_consumed_se = consumed_se,
         fcid_wasted_mean = wasted_mean,
         fcid_wasted_se = wasted_se,
         fcid_inedible_mean = inedible_mean,
         fcid_inedible_se = inedible_se)

# look at just inedible and wasted
enviro_final_dat %>% 
  filter(subgroup == 0) %>% 
  select(food, fcid_consumed_mean, fcid_wasted_mean, fcid_inedible_mean) %>% 
  head()

# something funky happening with inedible amts of meat
# update: this is correct, fcid weight basis for meat is boneless meat so inedible=0

```

### Environment/Social Impact Factors, Per 100 Grams and Per 1 DGA Serving

First, join with the conversion units.

```{r}

# merge with units
enviro_final_dat1 <- enviro_final_dat %>% 
  left_join(units, by = c("food" = "Food_group"))

enviro_final_dat1 %>% filter(is.na(Conversion_to_grams)) #good

```

Then, calculate the environment/social impact factors (i) per 100 grams and (ii) per DGA unit.

To calculate the impacts (e.g., GHG) per 100 grams, you divide the daily mean GHG of the food group by the daily amount of food consumed AND inedible, and then multiply by 100.

Note: The variable "ghg_mean_Consumed" represents the GHG impact of consumed and inedible food (because the impact of both the consumed and inedible amounts is included in the original datafield impact factors). Therefore, the denominator of the equation is the sum of consumed food AND inedible food.

```{r}

enviro_final_dat2 <- enviro_final_dat1 %>% 
  rowwise() %>% 
  mutate(#GHG
         GHGper100gram_consumed = ifelse(fcid_consumed_mean == 0, 0, (ghg_mean_Consumed / sum(fcid_consumed_mean, fcid_inedible_mean)) * 100),
         GHGperDGA_consumed = GHGper100gram_consumed * (Conversion_to_grams / 100),
         
         #CED
         CEDper100gram_consumed = ifelse(fcid_consumed_mean == 0, 0, (ced_mean_Consumed / sum(fcid_consumed_mean, fcid_inedible_mean)) * 100),
         CEDperDGA_consumed = CEDper100gram_consumed * (Conversion_to_grams / 100),
         
         #WATER
         WATERper100gram_consumed = ifelse(fcid_consumed_mean == 0, 0, (water_mean_Consumed / sum(fcid_consumed_mean, fcid_inedible_mean)) * 100),
         WATERperDGA_consumed = WATERper100gram_consumed * (Conversion_to_grams / 100),
         
         #BLUEWATER
         BLUEWATERper100gram_consumed = ifelse(fcid_consumed_mean == 0, 0, (bluewater_mean_Consumed / sum(fcid_consumed_mean, fcid_inedible_mean)) * 100),
         BLUEWATERperDGA_consumed = BLUEWATERper100gram_consumed * (Conversion_to_grams / 100),
         
         #FORCED LABOR
         FLper100gram_consumed = ifelse(fcid_consumed_mean == 0, 0, (fl_mean_Consumed / sum(fcid_consumed_mean, fcid_inedible_mean)) * 100),
         FLperDGA_consumed = FLper100gram_consumed * (Conversion_to_grams / 100)) %>% 
  select(-Conversion_to_grams)

```

## Finalize Impact Factors

### Merge Cost, Environment, and Social Impact Factors

Merge all of the impact factors into one dataset.

```{r}

# cost impact factors
cost_IFs <- cost_final_dat2 %>% 
  rename(costper100g = costper100gram_consumed,
         costperDGA = costperDGA_consumed) %>% 
  filter(!(food %in% c("babyfood", "coffee_tea", "other", "water")))

# environment impact factors
enviro_IFs <- enviro_final_dat2 %>% 
  filter(!(food %in% c("babyfood", "coffee_tea", "other", "water"))) %>% 
  rename(GHGperDGA = GHGperDGA_consumed,
         GHGper100g = GHGper100gram_consumed,
         
         CEDperDGA = CEDperDGA_consumed,
         CEDper100g = CEDper100gram_consumed,
         
         WATERperDGA = WATERperDGA_consumed,
         WATERper100g = WATERper100gram_consumed,
         
         BLUEWATERperDGA = BLUEWATERperDGA_consumed,
         BLUEWATERper100g = BLUEWATERper100gram_consumed,
         
         FLperDGA = FLperDGA_consumed,
         FLper100g = FLper100gram_consumed)

```

## Calculate Inedible and Wasted Proportions

### Environment/Social

The following proportions were calculated at the FCID-level:

- Wasted Proportion =  Mean Wasted Amount / Mean Edible Amount
- Inedible Proportion = Mean Inedible Amount / Mean Purchased Amount 

```{r}

enviro_IFs1 <- enviro_IFs %>% 
  rowwise() %>% 
  mutate(# coefficients
         wasted_coef_fcid = fcid_wasted_mean / fcid_consumed_mean,
         inedible_coef_fcid = fcid_inedible_mean / fcid_consumed_mean,
         
         # amounts
         fcid_purchased_mean = sum(fcid_consumed_mean, fcid_wasted_mean, fcid_inedible_mean),
         fcid_edible_mean = sum(fcid_consumed_mean, fcid_wasted_mean),
         
         # proportions
         wasted_prop_fcid = fcid_wasted_mean / fcid_edible_mean,
         inedible_prop_fcid = fcid_inedible_mean / fcid_purchased_mean,
         consumed_prop_fcid = fcid_consumed_mean / fcid_edible_mean,
         prop_sum = sum(wasted_prop_fcid, consumed_prop_fcid)) ## change to summing wasted/edible and consumed/edible so that this sums to one, and can use as a check

# is sum = 1?
table(enviro_IFs1$prop_sum, useNA = "always") #good

# create dataset with just IFs
enviro_IFs_sub <- enviro_IFs1 %>% 
  select(subgroup, food, 
         wasted_coef_fcid,
         wasted_prop_fcid,
         inedible_coef_fcid,
         inedible_prop_fcid,
         contains("perDGA"))

# look at a few foods
enviro_IFs_sub %>% filter(food == "gr_whole") %>% head()
enviro_IFs_sub %>% filter(food == "pf_redm_tot") %>% head()
enviro_IFs_sub %>% filter(food == "veg_dg") %>% head()

```

### Cost

The following proportions were calculated at the FNDDS-level:

- Wasted Proportion =  Mean Wasted Amount / Mean Edible Amount
- Inedible Proportion = Mean Inedible Amount / Mean Purchased Amount 

```{r}

cost_IFs1 <- cost_IFs %>% 
  rowwise() %>% 
  mutate(# coefficients
         wasted_coef_fndds = fndds_wasted_mean / fndds_consumed_mean,
         inedible_coef_fndds = fndds_inedible_mean / fndds_consumed_mean,
         
         # amounts
         fndds_purchased_mean = sum(fndds_consumed_mean, fndds_wasted_mean, fndds_inedible_mean),
         fndds_edible_mean = sum(fndds_consumed_mean, fndds_wasted_mean),
         
         # proportions
         wasted_prop_fndds = ifelse(fndds_edible_mean == 0, 0, fndds_wasted_mean / fndds_edible_mean), 
         inedible_prop_fndds = ifelse(fndds_purchased_mean == 0, 0, fndds_inedible_mean / fndds_purchased_mean),
         consumed_prop_fndds = ifelse(fndds_edible_mean == 0, 0, fndds_consumed_mean / fndds_edible_mean),
         prop_sum_fndds = sum(wasted_prop_fndds, consumed_prop_fndds))  ##c hange to summing wasted/edible and consumed/edible so that this sums to one, and can use as a check

table(cost_IFs1$prop_sum_fndds, useNA = "always")

# check
cost_IFs1 %>% 
  filter(wasted_prop_fndds == 0) %>% head()

# subset
cost_IFs_sub <- cost_IFs1 %>% 
  select(subgroup, food, food_type, mixed_dish, 
         wasted_coef_fndds, wasted_prop_fndds,
         inedible_coef_fndds, inedible_prop_fndds,
         costperDGA) 

# check a few
cost_IFs_sub %>% filter(food == "gr_whole") %>% head()
cost_IFs_sub %>% filter(food == "pf_redm") %>% head()

```

## Export Data

```{r}

# cost IFs
write_csv(cost_IFs_sub, paste("data_inputs/IMPACT_FACTORS/output_data/Impacts_cost_", my_date, ".csv", sep=""))

# enviro IFs
write_csv(enviro_IFs_sub, paste("data_inputs/IMPACT_FACTORS/output_data/Impacts_enviro_", my_date, ".csv", sep=""))

```

